The effects of probabilistic context inference on motor adaptation

### 2022-09-04 adaptation/article/article.tex
Plos One submission ready

\abstract{Abstract}
Humans have been shown to adapt their movements when sudden or gradual changes to the dynamics of the environment are introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Humans are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. Recently, the COIN computational model for motor adaptation and context inference was introduced, which contains components for context inference and Bayesian motor adaptation. This model was used to show the effects of context inference on learning rates across different experiments. We expanded on that work by using the COIN model to show that the effects of context inference on motor adaptation and control go even further than previously shown. Here, we used the COIN model to simulate classical motor adaptation experiments from previous works and showed that context inference, and how it is affected by the presence and reliability of feedback, effect a host of behavioral phenomena that had so far required multiple hypothesized mechanisms, lacking a unified explanation. Concretely, we show that the reliability of direct contextual information, as well as noisy sensory feedback, typical of many experiments, effect measurable changes in switching-task behavior, as well as in action selection, that stem directly from probabilistic context inference.

@@

In this work, we use the COIN model to show that the process of context inference underlies more behavioral phenomena than previously shown. We focused on the effects of uncertain contextual information on switching behavior, especially during so-called error-clamp trials, in which errors are forced to zero by experimenters. More specifically, we focused on the effects of perceptual noise, as well as feedback modalities, in context inference, which in turn affects behavior in ways that can be directly measured. We show that through context inference, switching behavior can display three main effects that have been previously attributed to hypothesized ad-hoc mechanisms: (1) The size of an adaptation dictates how quick and reliable switching between tasks is \cite{Oh_Minimizing_2019,Kim_Neural_2015}, which we explain in terms of the effects of perceptual noise on context inference. (2) Previously-learned adaptations can interfere with switching behavior \citep{Davidson_Scaling_2004}, which we explain in terms of uncertain context inference. (3) Training history (i.e. which adaptations have been learned and for how long) affects switching during error-clamp trials \citep{Vaswani_Decay_2013}, which we also attribute to uncertain context inference. To do this, we used the COIN model to simulate the experimental setups and the decision-making agents (i.e. participants) during those experiments.

With these combined simulations and their qualitative comparison to the experimental phenomena outlined, we provide further evidence that context inference may be a single coherent and mechanistic account that underlies experimentally well-established motor adaption and history effects under changing contexts.

\section{Results}
Using the COIN model \citep{Heald_Contextual_2021}, we simulated representative experiments from a number of experimental studies on motor adaptation to illustrate how this model explains different experimental findings using the dynamics of context inference. We will present these simulations alongside the experimental results from the representative studies and discuss in detail how context inference explains the experimental phenomena.

Before presenting these results, we briefly describe the COIN model, leaving a more thorough explanation for the methods section. Additionally, we present simulations using the model that show the effects of contextual cues and perceptual noise on context inference, which pave the way for the simulated experiments that we show later on.

@@

The COIN model contains, additionally to these three main components, components for learning new contexts (i.e. inferring the existence of a new context that had not been previously encountered by the agent), as well as the ability to infer subject-specific parameters such as a participant's assumed transition probabilities between contexts, which can differ from the real, hidden transition probabilities. In this work, we focused on switching behavior between previously-learned contexts, as well as in the perceptual aspect of context inference. More specifically, we simulated previous experiments and focused on the way the COIN model qualitatively explains experimental phenomena observed in these experiments. To do this, we fixed both the transition probabilities between contexts and the number of contexts based on the real values during the experiments, which were made available to participants. Fixing these values, in turn, allowed us to simplify the generative model, internal to the agent, using conjugate priors, with which the agent can do exact Bayesian inference. The generative model we used here can be seen in \fref{fig:model}A, including the priors for both context inference and motor adaptation. For more details on these choices, see the Methods section.

\begin{figure}
\centering
\includegraphics[]{./figures/figure_1.png}
\caption{Schematic representation of the model and illustrative simulations. (A) Generative model. Clear circles represent the observables, i.e. motor commands $u_t$ and direct observations $y_t$ (e.g. cursor position). The true context $\zeta_t$ is not directly observable, but influences $y_t$. The dark rectangles represent the prior distributions for the inferred adaptation level $x_t$ (Normal-Gamma distribution) and the current context $\zeta_t$ (discrete distribution with known $\pi$; see Methods). At every trial, the context is inferred, then motor adaptation is carried out and finally a motor command is issued; the flow of this process is indicated by the gray arrow in the background, while black arrows show the direction of information flow. (B) Simulations obtained with the model in (A), using a simulated experimental setup similar to that by \cite{Davidson_Scaling_2004}, in which the context changes at trial 20 (vertical, dashed lines). A total of 3x3 experiments were simulated, with low, medium and high levels of both cue uncertainty and observation noise. Each plot represents context inference $p(ctx)$ for one specific level of cue uncertainty and observation noise. The y-axis represents the posterior probability of each context $p(\zeta_t = j)$; the black line represents the baseline context (i.e. no adaptation), the green line represents the only adaptation to be learned during the simulated experiment. The shaded areas represent the standard deviation around the mean, obtained across 50 simulated participants.}
\label{fig:model}
\end{figure}

\subsection{Contextual cues and feedback}
The behavioral phenomena which are the focus of this work can be explained as arising from the effects of contextual cues and sensory feedback provided to participants during the experiment. To illustrate these effects in a simple example, we first simulated a generic motor adaptation experiment similar to those performed by \cite{Davidson_Scaling_2004}, in which participants must adapt to a visuomotor rotation of the cursor on a screen in relation to the movement of a pointing device.

The key to an intuitive understanding of the results presented below, is to observe what happens when the presence and reliability of contextual cues is varied, as well as the perceptual noise added to the cursor. In \fref{fig:model}B, a 3x3 grid of results can be seen, where each simulation in this grid is a combination of low, medium or high contextual cue reliability (where low reliability is equivalent to presenting no cues), and low, medium or high perceptual noise.

The first column of \fref{fig:model} shows that, in the presence of reliable contextual cues, context inference is accurate, certain and fast to switch. However, as contextual cues become less reliable, switching between known contexts becomes slower, as seen in the posterior probabilities over contexts at and after trial 20 in the second and third columns. Furthermore, as perceptual noise increases, switching becomes not only slower, but also more uncertain, with individual agents incorrectly missing the switch entirely.

Additionally, the motor responses (shown in purple) can be seen to depend not only on the current estimate of the rotation angle (shown as the green line), but also the uncertainty over the current context. This is most evident in the high cue uncertainty and observation noise panel, in which the response varies substantially from the estimate of the angle (i.e. the purple line deviates from the green line), even when the real context has the highest estimated probability.

@@

For clarity, we first introduce necessary terminology that is typically used in experimental studies. As an example, we will use a typical motor adaptation task in which participants have to make reaching movements while holding the handle of a mechanical arm that exerts a curl force on the participant's hand. Depending on the trial, the mechanical arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context O as that in which the mechanical arm exerts no force. Contexts A and B can be defined as those with clockwise or counter-clockwise, respectively. Abusing notation, a usual statement is that $B = (-A)$, as the forces have the same magnitude but point in opposite directions. Similarly, one can define context A/2, with the same direction of adaptation as A, but half the magnitude. Finally, many experiments include a block of error-clamp trials at the end of the experiment, in which the mechanical arm forces the participant to make straight-line movements; we represent these with the letter E.

With this terminology, a typical experiment \cite[e.g.][]{Ethier_Spontaneous_2008} would have a block structure of O-A-B-E, or O-A-(-A)-E, which means that the participant goes through a block of trials with no external force applied (O), a number of trials with a clockwise curl force (A), a block with counter-clockwise forces (B), and finally a block with error-clamp trials (E). With repeated contexts \citep[e.g.][]{Oh_Minimizing_2019}, an experiment can be described as $O_1$-$A_1$-$O_2$-\ldots .

\subsubsection{Savings and slow/fast switching}
The term 'savings' refer to the ability to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in humans \citep{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an O-A-O-A experiment, for example, savings would express themselves in the second A block in the form of a much higher adaptation rate than that observed during the first A block. The related concept of quick de-adaptation occurs in A-O transitions, where participants switch back to baseline without having to re-learn it.

In this section, we discuss savings and de-adaptation in terms of switching between contexts. We show that through context inference and how it is affected by contextual cues and observation noise, savings are not immediate, but a relatively fast process that reflects context inference.

@@

In \fref{fig:oh-2019}A we show the results of simulations with the model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts in the center panel, i.e. when the solid gray line (representing the true context) switches between 40, 0 and -40; in both panels (second and third), the thick black line, representing the agent's adaptation, quickly follows these switches.

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Savings and de-adaptation. Data from our simulations (first and second columns) compared to data adapted from figure 2A by \cite{Kim_Neural_2015} and figure 4A from \cite{Oh_Minimizing_2019} (last column). Experimental data was, in all three experiments, averaged across all participants; in our simulations, a single simulation is shown. In the first column, the simulated context over inference is represented by the posterior probabilities over all available contexts. Each color (black, blue, green) represents a different context, with black always representing the baseline (i.e. no adaptation). Vertical, dashed lines represent switches in the real context. In the second and third columns, the black line represents the adaptation (i.e. response) displayed by the agent as a function of trial number. The thin gray lines represent the optimal adaptation, i.e. the size of the true visuomotor rotation during the task. In all panels, the x-axis represents trial number, from 0 to 300. (A) Experiment by \cite{Kim_Neural_2015} with two visuomotor rotations (-40 and 40 degrees), in addition to baseline, with all possible transitions between A, (-A) and O. Participants must adapt to the rotations during shooting movements, and the participant's response can be understood as the rotation of the participants' shooting movements to compensate for the task's rotation. For clarity, both experimental and simulated results are shown only up to trial 300, of the original 600. (B) An experiment by \cite{Oh_Minimizing_2019} similar to (A), but the contexts were not cued and only one context, with an adaptation of 20 degrees, is introduced. O-A and A-O transitions (i.e. adaptation and de-adaptation) can be observed. Blue and black lines are as in A. (C) Same experiment as (B) but with a 10 degree adaptation.}
\label{fig:oh-2019}
\end{figure}

Critically, in \fref{fig:oh-2019}A it can be seen for the first four context switches (until about trial 250) that participants had not yet completely adapted to the rotation, as evidenced by their responses not being on par with the true rotation angle. This undershooting of responses (i.e. not doing the 40 degree rotation) happens despite participants being able to immediately and with high certainty identify the true context, as shown in the first-column panel. These results are similar to those shown by \cite{Imamizu_Explicit_2007}, where sensory cues of varying reliability effected immediate or slow contextual switches.

To expand on these results, we now turn to feedback and its effects on switching behavior. \cite{Oh_Minimizing_2019} performed two partially-cued experiments with a visual rotation of 20 and 10 degrees, respectively. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, alongside simulations with the COIN model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A$. In subsequent $A - O$ transitions, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in their responses (black line in the last two columns of \fref{fig:oh-2019}B) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation lagging behind the switches in the real context, and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained in our simulations in terms of the size of the adaptation in relation to observation noise: as the adaptation is smaller (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

Note that the results from our simulations from \cite{Oh_Minimizing_2019} and from \cite{Kim_Neural_2015} include both savings (O-A transitions) and de-adaptation (A-O transitions), both of which display the same characteristics and are explained by the same mechanism.

\subsubsection{Uncertainty over contexts affects action selection}
As with learning, action selection is affected by context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is (see \eref{eqn:dist-comm}).

Experimental evidence supporting this view can be found in experiments with context switching. For example, \cite{Davidson_Scaling_2004} reported a curl-force experiment in which participants had to switch from 3A to A in one group, with a block sequence A-3A-A-3A, and from -A to A in another group, with a block sequence A-(-A)-A-(-A). After A and 3A (or -A in the other group) had been learned in the first two blocks, the authors found that the switch from 3A to A was faster than that from -A to A. The authors interpreted this as evidence that switching between adaptations happens more quickly if it is in the same direction as the current adaptation (e.g. both counter-clockwise), and more slowly if they are in the opposite direction (e.g. clockwise to counter-clockwise).

Under the COIN model, the asymmetry is caused by the existence of the baseline context, which has a non-zero probability $p(\zeta_O | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $\zeta_O$ becomes more likely (given that $\zeta_{3A}$ has been ruled out). Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but in the second group, it slows down the switch because $A > 0 > -A$.

We simulated data with the model fitting the experimental structure in \cite{Davidson_Scaling_2004}. We show the results in \fref{fig:davidson-2004}A, alongside the experimental results from \cite{Davidson_Scaling_2004}. It can be seen that the agent exposed to the -A context shifts back to A more slowly than the one exposed to 3A, qualitatively reproducing the data from \cite{Davidson_Scaling_2004}.

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Motor error when switching back to a previously-learned adaptation. (A) Experimental results from \cite{Davidson_Scaling_2004} and simulations with the COIN model are shown. In the first column, each panel represents context inference for one group of participants (top: group A; bottom: group 3A), with each line representing the posterior probability of a context (black for the baseline O). The second column represents the error made by our simulated agent after returning to the previously-learned context, with blue and yellow representing groups -A and 3A, respectively. The last panel represents the same data, from the \cite{Davidson_Scaling_2004} experiment. All panels share the x-axis, representing the number of trials elapsed since the switch to the new context. All simulations were executed eight times per group, following the number of participants in the experiments; shaded areas represent the standard deviation across these eight simulations. (B) Simulated results for an experiment similar to \cite{Davidson_Scaling_2004}, but changing the contexts seen by the two groups from -A to -2A for the first group, and from 3A to 4A for the second group. The panels follow the same structure as (A), without the last panel for experimental results. (C) Simulations for an experiment in which the baseline context has been removed altogether.}
\label{fig:davidson-2004}
\end{figure}

To confirm our hypothesis, we simulated variants of the experiments in which the COIN model predicts that the difference between groups disappears. First, in \fref{fig:davidson-2004}B, we simulated an experiment in which the contexts have more extreme adaptations, making them more different from baseline than in the \cite{Davidson_Scaling_2004} experiments. To do this, one group adapts in a $O-A-(-2A)$ paradigm, while the other group adapts in a $0-A-4A$ paradigm. As in the original experiment, the second contexts ($-2A$ for one group, $4A$ for the other) are equally spaced from the first context. However, given the larger distance from baseline, the baseline context has a much lower attribution when the switch occurs compared to the original experiment. This change makes both simulated groups infer the correct context almost equally quickly, making the difference between their errors much smaller compared to the original experiment. Furthermore, we simulated an experiment with an identical structure to that of \cite{Davidson_Scaling_2004}, but eliminated the baseline context from the agent. The results can be seen in \fref{fig:davidson-2004}C, where the switches between contexts are made identically by the two groups. We discuss possible implementations of such an experiment in the Discussion section, alongside with other predictions made by the model.

\subsubsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Participants' behavior is consistent with a previously-encountered context (called spontaneous recover in O-A-B-E experiments, where behavior is consistent with A); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline, which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of adaptation during the first phase, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the response vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.

>>>>

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden or gradual change to the dynamics of the environment are introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Humans are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, affecting the switch between known adaptations. Recently, computational models for motor adaptation and context inference have been introduced, which contain components for context inference and Bayesian motor adaptation. These models were used to show the effects of context inference on learning rates across different experiments. We expanded on these works by using a simplificed version of the recently-introduced COIN model to show that the effects of context inference on motor adaptation and control go even further than previously shown. Here, we used this model to simulate classical motor adaptation experiments from previous works and showed that context inference, and how it is affected by the presence and reliability of feedback, effect a host of behavioral phenomena that had so far required multiple hypothesized mechanisms, lacking a unified explanation. Concretely, we show that the reliability of direct contextual information, as well as noisy sensory feedback, typical of many experiments, effect measurable changes in switching-task behavior, as well as in action selection, that stem directly from probabilistic context inference.

@@

In this work show that the process of context inference underlies more behavioral phenomena than previously shown. To do this, we used a minimal model for motor adaptation that includes context inference, which we derived by simplifying the COIN model (henceforth called sCOIN). We focused on the effects of uncertain contextual information on switching behavior, especially during so-called error-clamp trials, in which errors are forced to zero by experimenters. More specifically, we focused on the effects of perceptual noise, as well as feedback modalities, in context inference, which in turn affects behavior in ways that can be directly measured. We show that through context inference, switching behavior can display three main effects that have been previously attributed to hypothesized ad-hoc mechanisms: (1) The size of an adaptation dictates how quick and reliable switching between tasks is \citep{Oh_Minimizing_2019,Kim_Neural_2015}, which we explain in terms of the effects of perceptual noise on context inference. (2) Previously-learned adaptations can interfere with switching behavior \citep{Davidson_Scaling_2004}, which we explain in terms of uncertain context inference. (3) Training history (i.e. which adaptations have been learned and for how long) affects switching during error-clamp trials \citep{Vaswani_Decay_2013}, which we also attribute to uncertain context inference. To do this, we used the sCOIN model to simulate the experimental setups and the decision-making agents (i.e. participants) during those experiments.

Importantly, the goal of this work is not to introduce a new model for contextual motor learning, but to use the existing ideas of the COIN model to show that context inference can explain more experimental phenomena than those explored by \cite{Heald_Contextual_2021}.

With these combined simulations and the qualitative comparison to the experimental phenomena outlined above, we provide further evidence that context inference is a single, coherent and mechanistic account that underlies experimentally well-established motor adaption and history effects under changing contexts.

\section{Results}
Using the sCOIN model, we simulated representative experiments from a number of experimental studies on motor adaptation to illustrate how this model explains different experimental findings using the dynamics of context inference. We will present these simulations alongside the experimental results from the representative studies and discuss in detail how context inference explains the experimental phenomena.

Before presenting these results, we briefly describe the COIN model and the simplifications that led to the sCOIN version used in simulations. We leave a more thorough explanation of the models for the methods section. Additionally, we present simulations using the sCOIN model that show the effects of contextual cues and perceptual noise on context inference, which pave the way for the simulated experiments that we show later on.

@@

The COIN model contains, additionally to these three main components, components for learning new contexts (i.e. inferring the existence of a new context that had not been previously encountered by the agent), as well as the ability to infer subject-specific parameters such as a participant's assumed transition probabilities between contexts, which can differ from the real, hidden transition probabilities. Because we seeked to focus on switching behavior between previously-learned contexts, as well as in the perceptual aspect of context inference, we chose to fix the participant-inferred transition probabilities between contexts, as well as the total number of contexts; in our simulations, we assume that participants have already inferred the real values of these quantities. Because we mainly focus on switching behavior, as well as error-clamp trials (both of which involve alread-learned adaptations), these simplifications to the model have minimal effects on our results. See Methods for more details on the sCOIN model, as well as the differences between COIN and sCOIN.

By fixing the aforementioned values, the sCOIN model has a simpler generative model which allows the agent to perform exact Bayesian inference for motor adaptation. The inference process can be seen in \fref{fig:model}A, including the priors for both context inference and motor adaptation. For more details on these choices, see the Methods section.

\begin{figure}
\centering
\includegraphics[]{./figures/figure_1.png}
\caption{Schematic representation of the model and illustrative simulations. (A) Inference done by the model at every trial. Clear circles represent the observables, i.e. motor commands $u_t$ and direct observations $y_t$ (e.g. cursor position). The true context $\zeta_t$ is not directly observable, but influences $y_t$. The dark rectangles represent the prior distributions for the inferred adaptation level $x_t$ (Normal-Gamma distribution) and the current context $\zeta_t$ (discrete distribution with known $\pi$; see Methods). At every trial, the context is inferred, then motor adaptation is carried out and finally a motor command is issued; the flow of this process is indicated by the gray arrow in the background, while black arrows show the direction of information flow. (B) Simulations obtained with the model in (A), using a simulated experimental setup similar to that by \cite{Davidson_Scaling_2004}, in which the context changes at trial 20 (vertical, dashed lines). A total of 2x2 experiments were simulated, with low and high levels of both cue uncertainty and observation noise. In the left panel, the states inferred by the model for each of the 2x2 simulated experiments, where each color represents one experiment, with the same colors as the panels on the right. On the right, each plot represents context inference $p(ctx)$ for one specific level of cue uncertainty and observation noise. The y-axis represents the posterior probability of each context $p(\zeta_t = j)$; the black line represents the baseline context (i.e. no adaptation), the colored line (with the same colors as the panel on the left) represents the only adaptation to be learned during the simulated experiment. The shaded areas represent the standard deviation around the mean, obtained across 50 simulated participants.}
\label{fig:model}
\end{figure}

\subsection{Contextual cues and feedback}
The behavioral phenomena which are the focus of this work can be explained as arising from the effects of contextual cues and sensory feedback provided to participants during the experiment. To illustrate these effects in a simple example, we first simulated a generic motor adaptation experiment similar to those performed by \cite{Davidson_Scaling_2004}, in which participants must adapt to the curl forces exerted by a mechanical arm on the participants' hands as they perform pointing movements.

The key to an intuitive understanding of the results presented below is to observe what happens when the presence and reliability of contextual cues is varied, as well as the perceptual noise on the position of the hand. In \fref{fig:model}B, a 2x2 grid of results is shown: each simulation in this grid is a combination of low or high contextual cue uncertainty (where high uncertainty is equivalent to presenting no cues), and low or high perceptual noise (representing how well participants can detect deviations from the straight-line movements).

\fref{fig:model}B shows that, in the presence of reliable contextual cues, context inference is accurate, certain and fast to switch. However, as contextual cues become less reliable, switching between known contexts becomes slower, as seen in the posterior probabilities over contexts at and after trial 20 in the last column. Furthermore, as perceptual noise increases, switching becomes not only slower, but also more uncertain, with individual agents incorrectly missing the switch entirely. While cue uncertainty and observation noise have an effect on the motor adaptation process, as seen on the left-most panel in \fref{fig:model}B, in all simulations the hidden state (i.e. the force exerted by the mechanical arm, $x_{j,t}$) is quickly inferred.

@@

For clarity, we first introduce necessary terminology that is typically used in experimental studies. As an example, we will use a typical motor adaptation task in which participants have to make reaching movements while holding the handle of a mechanical arm that exerts a curl force on the participant's hand. Depending on the trial, the mechanical arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context $O$ as that in which the mechanical arm exerts no force. Contexts $A$ and $B$ can be defined as those with clockwise and counter-clockwise forces, respectively. Abusing notation, a usual statement is that $B = (-A)$, as the forces have the same magnitude but point in opposite directions. Similarly, one can define context $A/2$, with the same direction of adaptation as $A$, but half the magnitude. Finally, many experiments include a block of error-clamp trials at the end of the experiment, in which the mechanical arm forces the participant to make straight-line movements; we represent these with the letter $E$.

With this terminology, a typical experiment \cite[e.g.][]{Ethier_Spontaneous_2008} would have a block structure of $O-A-B-E$, or $O-A-(-A)-E$, which means that the participant goes through a block of trials with no external force applied ($O$), a number of trials with a clockwise curl force ($A$), a block with counter-clockwise forces ($B$), and finally a block with error-clamp trials ($E$). With repeated contexts \citep[e.g.][]{Oh_Minimizing_2019}, an experiment can be described as $O_1-A_1-O_2-\ldots$.

\subsubsection{Cue- and sensory feedback uncertainty affects switching behavior}
The term 'savings' refer to the ability to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in humans \citep{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an $O-A-O-A$ experiment, for example, savings would express themselves in the second A block in the form of a much higher adaptation rate than that observed during the first $A$ block. The related concept of quick de-adaptation occurs in $A-O$ transitions, where participants switch back to baseline without having to re-learn it.

In this section, we discuss savings in terms of switching between contexts. We show that through context inference and how it is affected by contextual cues and observation noise, savings are not immediate, but a relatively fast process that reflects context inference. In particular, we show that the manifestations of savings on behavior are mediated by context inference, which could mask the presence of savings in cases where observations do not unequivocally identify a context.

@@

In \fref{fig:oh-2019}A we show the results of simulations with the model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts in the center panel, i.e. when the solid gray line (representing the true context) switches between 40, 0 and -40; in the two right-most panels, the thick black line, representing the agent's adaptation, quickly follows these switches.

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Switching between learned adaptations gated by context inference. Data from our simulations (first three columns) compared to data adapted from figure 2A by \cite{Kim_Neural_2015} and figure 4A from \cite{Oh_Minimizing_2019} (last column). Experimental and simulated data was, in all three experiments, averaged across all participants. In the first column, the simulated context inference is represented by the posterior probabilities over all available contexts. Each color (black, blue, green) represents a different context, with black always representing the baseline (i.e. no adaptation). Vertical, dashed lines represent switches in the real context. The inferred state (angle of the visuomotor rotation) is shown in the second column, with the same colors as in the first column. In the third and fourth columns, the black line represents the adaptation (i.e. response) displayed by the agent as a function of trial number. The thin gray lines represent the optimal adaptation, i.e. the size of the true visuomotor rotation during the task. (A) Experiment by \cite{Kim_Neural_2015}. Both experimental and simulated results are shown only up to trial 300, of the original 600. (B) An experiment by \cite{Oh_Minimizing_2019}.  Blue and black lines are as in A. (C) Same experiment as (B) but with a 10 degree adaptation.}
\label{fig:oh-2019}
\end{figure}

In the second panel of \fref{fig:oh-2019}A, we show the inferred state (i.e. the inferred angle of rotation). Critically, it can be seen for the first four context switches (until about trial 200) that participants had not yet completely adapted to the rotation, as also evidenced by their responses not being on par with the true rotation angle. This undershooting of responses (i.e. not doing the 40 degree rotation) happens despite participants being able to immediately and with high certainty identify the true context, as shown in the first-column panel. These results are similar to those shown by \cite{Imamizu_Explicit_2007}, where sensory cues of varying reliability effected immediate or slow contextual switches.

To expand on these results, we now turn to feedback and its effects on switching behavior. \cite{Oh_Minimizing_2019} performed two partially-cued experiments with a visuomotor rotation of 20 and 10 degrees, respectively. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, alongside simulations with the sCOIN model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A$. In subsequent context transitions, participants showed immediate switching (with a one-trial lag) between $A$ and $O$ (both ways), which can be seen in their responses (black line in the last two columns of \fref{fig:oh-2019}B) closely following the switches in the true rotation. In the second experiment (\fref{fig:oh-2019}C), context switching happens more slowly, with adaptation lagging behind the switches in the real context, and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained in our simulations in terms of the size of the adaptation in relation to observation noise: as the adaptation is smaller (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

In contrast, \cite{Oh_Minimizing_2019} explained the results of their second experiment by positing that when adaptations were small, participants did not identify this as a new context and opted instead for a modification of their baseline model (i.e. how they move normally). Under this single-context explanation, however, savings do not exist, and adaptations need to be learned anew ever time a context changes. On the other hand, savings are present under the sCOIN model, their manifestations being diminished by the slower context inference. \cite{Oh_Minimizing_2019} analyzed savings during their second experiment and found that savings do exist, although greatly diminished compared to the first experiment. This is in favor of the dual-context model, as we present it here. \cite{Oh_Minimizing_2019} showed a slow decay during error-clamp trials in their experiments, which they considered evidence for the single-context account of the second experiment. However, as we show below, this can also be explained in a dual-context model as an effect of slow context inference.

\subsubsection{Uncertainty over contexts affects action selection}
As with learning, we show in this section that action selection is affected by context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is (see \eref{eqn:dist-comm}).

Experimental evidence supporting this view can be found in experiments with context switching. For example, \cite{Davidson_Scaling_2004} reported a curl-force experiment in which participants had to switch between $A$ and $3A$ in one group, with a block sequence $A-3A-A-3A$, and from $A$ to $-A$ in another group, with a block sequence $A-(-A)-A-(-A)$. After $A$ and $3A$ (or $-A$ in the other group) had been learned in the first two blocks, the authors found that the switch from $3A$ to $A$ was faster than that from $-A$ to $A$. The authors interpreted this as evidence that switching between adaptations happens more quickly if it is in the same direction as the current adaptation (e.g. both counter-clockwise), and more slowly if they are in the opposite direction (e.g. clockwise to counter-clockwise).

Under the sCOIN model, the asymmetry is caused by the existence of the baseline context, which has a non-zero probability $p(\zeta_O | s_t ...)$, as can be seen in \fref{fig:davidson-2004}A. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $\zeta_O$ becomes more likely (given that $\zeta_{3A}$ has been ruled out). Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but in the second group, it slows down the switch because $A > 0 > -A$ and the behavior lingers around $0$ until $p(\zeta_0|\ldots)$ drops back to zero.

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Motor error when switching back to a previously-learned adaptation. (A) Experimental results from \cite{Davidson_Scaling_2004} and simulations with the SCOIN model are shown. In the first column, each panel represents context inference for one group of participants (top: group A; bottom: group 3A), with each line representing the posterior probability of a context (black for the baseline O). The second column represents the error made by our simulated agent after returning to the previously-learned context, with blue and yellow representing groups -A and 3A, respectively. The last panel represents the same data, from the \cite{Davidson_Scaling_2004} experiment. All panels share the x-axis, representing the number of trials elapsed since the switch to the new context. All simulations were executed 32 times per group, to obtain a reliable mean; shaded areas represent the standard deviation across all simulations. (B) Simulated results for an experiment similar to \cite{Davidson_Scaling_2004}, but changing the contexts seen by the two groups from -A to -2A for the first group, and from 3A to 4A for the second group. The panels follow the same structure as (A), without the last panel for experimental results. (C) Simulations for an experiment in which the baseline context has been removed altogether.}
\label{fig:davidson-2004}
\end{figure}

To confirm this explanation, we simulated variants of the experiment in which the sCOIN model predicts that the difference between groups diminishes or disappears. First, in \fref{fig:davidson-2004}B, we simulated an experiment in which the contexts have more extreme adaptations, making them more different from baseline than in the \cite{Davidson_Scaling_2004} experiments. To do this, one group adapts in a $O-A-(-2A)$ paradigm, while the other group adapts in a $0-A-4A$ paradigm. As in the original experiment, the second contexts ($-2A$ for one group, $4A$ for the other) are equally spaced from the first context. However, given the larger distance from baseline, the baseline context has the same probability for both groups after the switch back to $A$. This change makes both simulated groups infer the correct context almost equally quickly, making the difference between their errors much smaller compared to the original experiment. Furthermore, we simulated an experiment with an identical structure to that of \cite{Davidson_Scaling_2004}, but eliminated the baseline context from the agent. The results can be seen in \fref{fig:davidson-2004}C, where the switches between contexts are made identically by the two groups.

\subsubsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Participants' behavior is consistent with a previously-encountered context (called spontaneous recover in $O-A-B-E$ experiments, where behavior is consistent with $A$); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline, which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of adaptation during the first phase, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the response vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.
